---
title: "Causal Forests for inspecting the heterogeneous effects of interventions on
  pro-climate behaviour"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

## Preprocessing

Notebook to run causal forest on our different interventions. This is our preparatory work.

```{r}
library(grf)
library(dplyr)
library(ggplot2)
library(tidyr)


df <- read.csv("clean2.csv")


df$Share[df$Share == 2] <- NA
df$Gender[df$Gender == 3] <- NA
df$Education.2[df$Education.2 == 2] <- NA
df$Income[df$Income == 9] <- NA

df <- na.omit(df)

df <- df %>%
  rename(
    Education = Education.2,
    Politics = Politics2_1,
    SES = MacArthur_SES,
    Country = country
  )

```

We can change this to the following different outcomes: `belief_avg, WEPT_max, policy_avg, Share`

```{r}
Y <- df$Share #change this as needed

```

Now we define our covariates as well as the different interventions utilised (includes control). We vectorise our categorical variables gender and country into ordinal integers.

```{r}

covar_raw <- c("Age", "Gender", "Education", "Politics",
               "Income", "SES", "Country")

W <- as.factor(df$condName) # Interventions



covars <- df[, covar_raw]

# Flag + mean/mode-impute missing values (keeps rows instead of dropping)
for (col in names(covars)) {
  if (is.numeric(covars[[col]])) {
    covars[[col]][is.na(covars[[col]])] <- mean(covars[[col]], na.rm = TRUE)
  } else {
    covars[[col]][is.na(covars[[col]]) | covars[[col]] == ""] <- "Missing"
  }
}
# Vectorize categorical covariates (Gender, country) into integer codes
covars$Gender[is.na(covars$Gender) | covars$Gender == ""] <- "Missing"
covars$Country[is.na(covars$Country) | covars$Country == ""] <- "Missing"

covars$Gender  <- as.integer(factor(covars$Gender))  - 1
covars$Country <- as.integer(factor(covars$Country)) - 1
X <- as.matrix(covars)


# X <- model.matrix(~ . -1, data = X)

```

## Training (And Evaluating) a Causal Forest

We fit a multi-arm causal forest to estimate the heterogeneous effects of different climate interventions on pro-climate behaviour. Since intervention effectiveness may vary substantially between individuals depending on their characteristics, causal forests estimates conditional average treatment effects (CATEs). This allows the effect of each intervention to vary as a function of observed participant characteristics. This enables us to identify not only which interventions are most effective overall, but also which interventions are most effective for specific subgroups. 

Participants were randomly assigned to a single intervention condition, allowing the model to compare differences in outcomes between individuals exposed to different treatments. Causal Forests allow us to identify not only which interventions are most effective overall, but also which interventions are most effective for specific subgroups. We use a multi-arm causal forest because the study includes multiple treatment conditions rather than a single binary intervention. The model estimates the expected outcome under each treatment condition and identifies how the relative effectiveness of interventions varies across individuals. Unlike conventional regression approaches, which require researchers to specify interaction terms a priori, causal forests use machine learning to detect complex and potentially non-linear relationships between participant characteristics and treatment effectiveness.


```{r}
forest <- multi_arm_causal_forest(
  X = X,
  Y = Y,
  W = W,
  num.trees = 1000,
  honesty = TRUE
)
```

We can evaluate the performance of our model using two checks

```{r}
set.seed(123)
train <- sample(
  1:nrow(X),
  0.8*nrow(X)
)
forest_train <- multi_arm_causal_forest(
  X = X[train,],
  Y = Y[train],
  W = W[train],
  num.trees = 1000
)
pred_test <- predict(
  forest_train,
  X[-train,]
)

############################################################
# Check 1: Covariate mean differences between train/test
# (sanity check that the split didn't accidentally create
# imbalanced samples - should all be small/non-significant
# since this is a random split)
############################################################
covariate_names <- colnames(X)

balance_check <- data.frame(
  covariate   = covariate_names,
  mean_train  = colMeans(X[train, , drop = FALSE], na.rm = TRUE),
  mean_test   = colMeans(X[-train, , drop = FALSE], na.rm = TRUE)
)
balance_check$mean_diff <- balance_check$mean_train - balance_check$mean_test

# Add a simple t-test p-value per covariate
balance_check$p_value <- sapply(covariate_names, function(cn) {
  t.test(X[train, cn], X[-train, cn])$p.value
})

print(balance_check)
cat("\nFlagged (p < 0.05, unexpected for a random split):\n")
print(balance_check[balance_check$p_value < 0.05, ])

############################################################
# Check 2: MSE between predicted and actual treatment effects
#
# Individual-level CATEs aren't directly observable (only one
# potential outcome is observed per person), so "actual" here
# is the OBSERVED average effect per arm in the test set
# (mean(Y | W = arm) - mean(Y | W = Control)), compared against
# the forest's PREDICTED average effect per arm on the same
# test units. This is the standard practical validity check.
############################################################
arm_names <- dimnames(pred_test$predictions)[[2]]
arm_names <- gsub(" - .*$", "", arm_names)
Y_test <- Y[-train]
W_test <- W[-train]

actual_effect <- sapply(arm_names, function(arm) {
  mean(Y_test[W_test == arm], na.rm = TRUE) -
    mean(Y_test[W_test == "Control"], na.rm = TRUE)
})

predicted_effect <- apply(pred_test$predictions[, , 1], 2, mean, na.rm = TRUE)

effect_comparison <- data.frame(
  arm              = arm_names,
  actual_effect    = actual_effect,
  predicted_effect = predicted_effect
)
effect_comparison$sq_error <- (effect_comparison$actual_effect -
                                  effect_comparison$predicted_effect)^2

print(effect_comparison)

mse_arm_level <- mean(effect_comparison$sq_error, na.rm = TRUE)
cat("\nArm-level MSE (predicted vs. actual average effect):", mse_arm_level, "\n")

```


We now compute the individual treatment effect and variable importance form the model


```{r}

tau <- predict(forest)$predictions # Individual Treatment effect

head(tau)

# Variable importance
importance <- variable_importance(forest)

importance_df <- data.frame(
  Variable = colnames(X),
  Importance = importance
)

importance_df <- importance_df %>%
  arrange(desc(Importance))

ggplot(importance_df,
       aes(x = reorder(Variable, Importance),
           y = Importance)) +
  geom_col() +
  coord_flip() +
  theme_minimal()
```

## How do the interventions effect subgroups of different ages?


```{r}

############################################################
# Subgroup analysis for Age (binned)
############################################################
library(dplyr)
library(tidyr)
library(ggplot2)

interventions <- dimnames(tau)[[2]]

cate_df <- data.frame(
  Age = df$Age,
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

# Bin Age into groups - adjust breaks to taste
cate_df <- cate_df %>%
  mutate(
    Age_bin = cut(
      Age,
      breaks = c(0, 18, 25, 35, 45, 55, 65, Inf),
      labels = c("<18", "18-24", "25-34", "35-44", "45-54", "55-64", "65+"),
      right = FALSE
    )
  )

cate_long <- cate_df %>%
  pivot_longer(
    cols = -c(Age, Age_bin),
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    !is.na(Age_bin),
    is.finite(Effect)
  )

# Summarise: mean + SE of CATE per bin, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, Age_bin) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = Age_bin, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = "Age group",
    y = "Estimated treatment effect",
    title = "Treatment effect by age group"
  )

```
## How do the interventions effect subgroups of different Education levels?

```{r}
ordinal_var <- "Education"  

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(ordvar),
    is.finite(Effect)
  ) %>%
  mutate(ordvar = factor(ordvar, ordered = TRUE))  # preserve level order

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = ordinal_var,
    y = "Estimated treatment effect",
    title = paste("Treatment effect by Education")
  )
```

## How do the interventions effect subgroups of different Income levels?

```{r}
ordinal_var <- "Income"   

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(ordvar),
    is.finite(Effect)
  ) %>%
  mutate(ordvar = factor(ordvar, ordered = TRUE))  # preserve level order

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = ordinal_var,
    y = "Estimated treatment effect",
    title = paste("Treatment effect by", ordinal_var)
  )
```


## How do the interventions effect subgroups on the Political spectrum?

```{r}
ordinal_var <- "Politics"

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

# Bin the 0-100 slider into a handful of ordered groups
cate_df <- cate_df %>%
  mutate(
    ordvar = cut(
      ordvar,
      breaks = c(-Inf, 20, 40, 60, 80, Inf),
      labels = c("1", "2", "3", "4", "5"),  # placeholder levels, relabeled below
      ordered_result = TRUE
    )
  )

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    !is.na(ordvar),
    is.finite(Effect)
  )

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  scale_x_discrete(
    labels = c("More liberal", "", "", "", "More conservative")
  ) +
  labs(
    x = "Political ideology",
    y = "Estimated treatment effect",
    title = "Treatment effect by political ideology"
  )
```





## How do the interventions effect subgroups on different standings in society (Macarther SES)?

```{r}
ordinal_var <- "SES"

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(ordvar),
    is.finite(Effect)
  ) %>%
  mutate(ordvar = factor(ordvar, ordered = TRUE))  # preserve level order

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = ordinal_var,
    y = "Estimated treatment effect",
    title = paste("Treatment effect by Subjective Socioeconomic Standing")
  )
```

## How do the interventions effect different genders?

```{r}

interventions <- dimnames(tau)[[2]]

cate_df <- data.frame(
  Gender = df$Gender,
  tau
)


cate_df <- cate_df %>%
  mutate(
    Gender = recode(
      as.character(Gender),
      "1" = "Male",
      "2" = "Female",
      "3" = "Prefer not to say",
      "4" = "Self-described"
    )
  )


colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -Gender,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    !is.na(Gender),
    is.finite(Effect)
  )

# Summarise: mean + SE of CATE per intervention, per gender
cate_summary <- cate_long %>%
  group_by(Intervention, Gender) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = Intervention, y = mean_effect, fill = factor(Gender))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  theme_minimal() +
  labs(
    x = "Intervention",
    y = "Estimated treatment effect",
    fill = "Gender",
    title = "Treatment effect by gender"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


