---
title: "Causal Forests for inspecting the heterogeneous effects of interventions on
  pro-climate behaviour"
output:
  pdf_document: default
  html_notebook: default
---

## Preprocessing

Notebook to run causal forest on our different interventions. This is our preparatory work.

```{r}
library(grf)
library(dplyr)
library(ggplot2)
library(tidyr)


df <- read.csv("clean2.csv")

```

We can change this to the following different outcomes: `belief_avg, WEPT_max, policy_avg, Share`

```{r}
Y <- df$policy_avg #change this as needed

```

Now we define our covariates as well as the different interventions utilised (includes control). We vectorise our categorical variables gender and country into ordinal integers.

```{r}

covar_raw <- c("Age", "Gender", "Education.2", "Politics2_1",
               "Income", "MacArthur_SES", "country")

W <- as.factor(df$condName) # Interventions



covars <- df[, covar_raw]

# Flag + mean/mode-impute missing values (keeps rows instead of dropping)
for (col in names(covars)) {
  if (is.numeric(covars[[col]])) {
    covars[[col]][is.na(covars[[col]])] <- mean(covars[[col]], na.rm = TRUE)
  } else {
    covars[[col]][is.na(covars[[col]]) | covars[[col]] == ""] <- "Missing"
  }
}
# Vectorize categorical covariates (Gender, country) into integer codes
covars$Gender[is.na(covars$Gender) | covars$Gender == ""] <- "Missing"
covars$country[is.na(covars$country) | covars$country == ""] <- "Missing"

covars$Gender  <- as.integer(factor(covars$Gender))  - 1
covars$country <- as.integer(factor(covars$country)) - 1
X <- as.matrix(covars)


# X <- model.matrix(~ . -1, data = X)

```

## Training (And Evaluating) a Causal Forest

We fit a multi-arm causal forest to estimate the heterogeneous effects of different climate interventions on pro-climate behaviour. Since intervention effectiveness may vary substantially between individuals depending on their characteristics, causal forests estimates conditional average treatment effects (CATEs). This allows the effect of each intervention to vary as a function of observed participant characteristics. This enables us to identify not only which interventions are most effective overall, but also which interventions are most effective for specific subgroups. 

Participants were randomly assigned to a single intervention condition, allowing the model to compare differences in outcomes between individuals exposed to different treatments. Causal Forests allow us to identify not only which interventions are most effective overall, but also which interventions are most effective for specific subgroups. We use a multi-arm causal forest because the study includes multiple treatment conditions rather than a single binary intervention. The model estimates the expected outcome under each treatment condition and identifies how the relative effectiveness of interventions varies across individuals. Unlike conventional regression approaches, which require researchers to specify interaction terms a priori, causal forests use machine learning to detect complex and potentially non-linear relationships between participant characteristics and treatment effectiveness.


```{r}
forest <- multi_arm_causal_forest(
  X = X,
  Y = Y,
  W = W,
  num.trees = 1000,
  honesty = TRUE
)
```

We can evaluate the performance of our model using two checks

```{r}
set.seed(123)

train <- sample(
  1:nrow(X),
  0.8*nrow(X)
)

forest_train <- multi_arm_causal_forest(
  X = X[train,],
  Y = Y[train],
  W = W[train],
  num.trees = 2000
)

pred_test <- predict(
  forest_train,
  X[-train,]
)
```


We now compute the individual treatment effect and variable importance form the model


```{r}

tau <- predict(forest)$predictions # Individual Treatment effect

head(tau)

# Variable importance
importance <- variable_importance(forest)

importance_df <- data.frame(
  Variable = colnames(X),
  Importance = importance
)

importance_df <- importance_df %>%
  arrange(desc(Importance))

ggplot(importance_df,
       aes(x = reorder(Variable, Importance),
           y = Importance)) +
  geom_col() +
  coord_flip() +
  theme_minimal()
```

## How do the interventions effect subgroups of different ages?


```{r}

interventions <- dimnames(tau)[[2]]
interventions

cate_df <- data.frame(
  Age = df$Age,
  tau
)

colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -Age,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(Age),
    is.finite(Effect)
  )

ggplot(cate_long, aes(Age, Effect, colour = Intervention)) +
  geom_smooth(
    method = "loess",
    se = FALSE
  ) +
  theme_minimal() +
  labs(
    x = "Age",
    y = "Estimated treatment effect"
  )

```
## How do the interventions effect subgroups of different Education levels?

```{r}
ordinal_var <- "Education.2"   # swap for "Income", "MacArthur_SES", etc.

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(ordvar),
    is.finite(Effect)
  ) %>%
  mutate(ordvar = factor(ordvar, ordered = TRUE))  # preserve level order

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = ordinal_var,
    y = "Estimated treatment effect",
    title = paste("Treatment effect by", ordinal_var)
  )
```

## How do the interventions effect subgroups of different Income levels?

```{r}
ordinal_var <- "Income"   

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(ordvar),
    is.finite(Effect)
  ) %>%
  mutate(ordvar = factor(ordvar, ordered = TRUE))  # preserve level order

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = ordinal_var,
    y = "Estimated treatment effect",
    title = paste("Treatment effect by", ordinal_var)
  )
```


## How do the interventions effect subgroups on the Political spectrum?

```{r}
ordinal_var <- "Politics2_1"

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(ordvar),
    is.finite(Effect)
  ) %>%
  mutate(ordvar = factor(ordvar, ordered = TRUE))  # preserve level order

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = ordinal_var,
    y = "Estimated treatment effect",
    title = paste("Treatment effect by", ordinal_var)
  )
```





## How do the interventions effect subgroups on different standings in society (Macarther SES)?

```{r}
ordinal_var <- "MacArthur_SES"

cate_df <- data.frame(
  ordvar = df[[ordinal_var]],
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -ordvar,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    is.finite(ordvar),
    is.finite(Effect)
  ) %>%
  mutate(ordvar = factor(ordvar, ordered = TRUE))  # preserve level order

# Summarise: mean + SE of CATE at each level, per intervention
cate_summary <- cate_long %>%
  group_by(Intervention, ordvar) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = ordvar, y = mean_effect, colour = Intervention, group = Intervention)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_pointrange(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.3),
    size = 0.3
  ) +
  theme_minimal() +
  labs(
    x = ordinal_var,
    y = "Estimated treatment effect",
    title = paste("Treatment effect by", ordinal_var)
  )
```

## How do the interventions effect different genders?

```{r}

interventions <- dimnames(tau)[[2]]

cate_df <- data.frame(
  Gender = df$Gender,
  tau
)
colnames(cate_df)[-1] <- gsub(" - ", "", interventions)

cate_long <- cate_df %>%
  pivot_longer(
    cols = -Gender,
    names_to = "Intervention",
    values_to = "Effect"
  ) %>%
  filter(
    !is.na(Gender),
    is.finite(Effect)
  )

# Summarise: mean + SE of CATE per intervention, per gender
cate_summary <- cate_long %>%
  group_by(Intervention, Gender) %>%
  summarise(
    mean_effect = mean(Effect),
    se = sd(Effect) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

ggplot(cate_summary, aes(x = Intervention, y = mean_effect, fill = factor(Gender))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_effect - 1.96 * se, ymax = mean_effect + 1.96 * se),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  theme_minimal() +
  labs(
    x = "Intervention",
    y = "Estimated treatment effect",
    fill = "Gender",
    title = "Treatment effect by gender"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


